Motion planning for assessing accessibility of protein tunnels

### 2017-09-09 paper.tex
text

%Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
%to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
%the random sample $\qrand$ is connected to the tree if the line segment  from $\qnear$ to $\qrand$ is collision-free. Otherwise, the retraction-step is performed.
%The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complexity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
Selective Retraction-based RRT~\cite{lee2012srrrt} avoid expensive growth in the open areas of the configuration space and focus
the retractions only to narrow passages.
Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on e.g. random vectors, obstacles vectors or medial axis in order to improve growth in the narrow passages.
In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed toward the medial axis.
%Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g. by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligands is modeled as a kinematic chain, which increases dimension of the configuration space.
To cope with many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} is employed as the basic planner.
RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., that are responsible for conformation changes) if they hinder the growth of the tree.

@@

In this paper, we propose a modification of the RRT planner for computing trajectories through protein tunnels.
The input is a set of tunnels computed by Voronoi diagrams and the task is to find trajectories along the given tunnels
leading from outside the protein to the active site.

In contrast to~\cite{cortes2010simulating}, which finds any exit pathway, we compute the trajectories only along each tunnel separately.
This is motivated by the practical needs of chemists that are interested in the traversability of a selected tunnel.
Computing trajectories along a single tunnel also brings advantages from the motion planning point of view.
It decreases the volume of the configuration space to be search and increases the relative volume of the narrow passages.
To further cope with the narrow passage problem, the random samples are not generated in the whole configuration space, but only (mainly) around
the tunnel in a user-defined distance.
Therefore, the tunnels serve as a guiding path for the configuration tree~\cite{vonasek2009rrt}.

As the input data are snapshots of the proteins without ligands, it can be expected, that the tunnels are narrow therefore it would be difficult
to find trajectories for all ligands.
In reality, the tunnels adapt to the ligands (and vice versa), so even narrow tunnels can serve as transportation path for ligands larger than
the tunnel bottleneck TODO CITACE??.
To emulate this behavior, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating,guieysse2008structure}.

The ligand flexibility is modeled using a predefined set of typical conformations, therefore the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
While classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
This metric measures distances between two configurations as the nearest distance of two atoms. 
This enables the metric to consider the shape of the ligand and it support retraction of the ligand to the narrow passages.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
The flexibility of a ligand is modeled using the set $\L$ of its conformations.
The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.

%or downloaded from a library~\cite{dunbrack}.
%We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

%Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
%To decrease the computational burden, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
%For each conformation $c\in \L$ an energy potential is also calculated.
%Let $\L_i$ denote the set of conformations that 
%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
%can be switched to the confomration $c_j \in C$.

@@

%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
To enable motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
A discrete set of scales is used, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., its geometric center), rotation around $x$, $y$, and $z$ axes,
index of the conformation $l\in \L$ and the actual scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various conformations.
The traversability of the tunnel should therefore be evaluated using various starting configurations $\QI$.
The initial configurations are have to be searched around the beginning of the tunnel.
To find a new initial configuration, a random sample $q$ is generated around $c_1$ in the distance $\RI$ (translation and rotation
 parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration, so $q$ is added to $\QI$.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$,  $K$ trajectories are created.
Each trajectory is computed using a modified RRT, which is introduced in the following section.

@@

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, the virtual goal is $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
%This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
This allows the tree to slightly detour from the tunnel, which improves performance around narrow tunnels.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
metric for the nearest-neighbor search.
The parameter $\rv$ influences the distribution of samples around the tunnel centerline. 
By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
of $\rv$ cause  exploration of paths around the tunnel.
We propose to set this parameter to the average tunnel width.

@@

The second proposed modification of RRT is the expansion procedure (Alg.~\ref{alg::expand}), which generates new collision-free nodes around $\qnear\in\T$.
For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qrand$ and tested for collisions.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
part is generated around $\qnear$.

@@

Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration
$q$ being tested.
This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}.
As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
The fast evaluation of this metric with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, can be achieved using KD-trees.
For each conformation $\l \in \L$ of the ligand, a KD-tree is build for the ligand at configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find a nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find nearest point towards $\qrand'$, which
is also the nearest neighbor of the ligand at $q$ to $\qrand$.

@@

The result of each planning trial is the tree $\T$ of collision-free configurations in which a path
between $\qinit$ (root of the tree) and $q'$ is found, where
 $q'$ is the nearest node in the tree towards $\qgoal$ (using 3D Euclidean metric). 
The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
The path is found in the tree even if the tree does not approach $\qgoal$ close enough.
Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.

@@

The proposed planner utilizes a set of predefined ligand's conformations.
As each of them is tested in the expansion step, this number of influences both runtime and the success rate.
By using more conformations, there is a bigger chance that at least one of them can expand the tree towards the random sample $\qrand$.
However, using large number of conformations also increase the time complexity of the planner.
Therefore, it is important so select a `good' set of conformations.
Depending on the ligand, some conformations are more packed i.e., the atoms in that conformations are close to each other, but 
conformations where the atoms form a longer molecule are also possible.

From motion planning point of view, both types of conformations are useful.
The `longer' conformations can better fit in a narrow passage, while the `packed' conformations can be better rotated at place 
(Fig.~\ref{fig::m040c}.
To select the suitable conformations, a simple technique is proposed.  
For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
The conformations are sorted according to this deviation and then the desired number of conformations are selected from this sorted list uniformly.

\begin{figure}
\centering
%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
\includegraphics[width=0.15\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.12\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 Examples of `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

The proposed approach was used to analyze tunnel traversability towards to active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
Molecular dynamics of the protein was computed using Amber tool.
%100 frames of the dynamics were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
In each frame, tunnels for the spherical probe 0.9~\AA were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck of the tunnels are $1.48$~\AA and $1.2$~\AA for the first, resp. the second tunnel.
%The second tunnel is more dynamic, i.e., it fluctuatest more

@@

Two motion planners were utilized: the proposed planner (Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paperm, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=10,000$, $m=100$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
The Retraction-based RRT was implemented using 6D Euclidean metric.
The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configurations $\qnew$ that approaches $\qrand$.
The retraction step was realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.1$~\AA\ using 20 samples.
A collision-free sample that maximally approached $\qrand$ was selected and added to the tree, and the retraction continued from this sample.
The retraction step was repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per conformation, while \RA\ extends the tree only by one configuration.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for nearest-neighbor search and the time to solve these queries is significantly less than computing collision detection.
Parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is same.
For each conformation and tested scale, the \RA\ methods attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.

@@

The results are shown in Tab.~\ref{tab::one} which shows percentage of MD frames where the sphere traversed the whole tunnel 
(i.e., when it reached the active site).
As the bottlenecks of the tunnels in each frame is known, we can compute the percentage of frames that have the bottleneck larger
than the tested sphere radius.
The first line in the Tab.~\ref{tab::one} therefore show the maximum percentage of traversable frames, while the second
line (denote Planners) shows how both planners performed in this scenario.
We can conclude, that the performance of both methods is similar in the case of first tunnel.
In the case of the second tunnel, \RA\ was able to find trajectory in more frames than \RB.

@@

The aim of this experiment is to verify the procedure to select ligand conformations for the motion planning.
The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations where available and for the ligand
m040 (1,5-Dichloropentane) with $7560$ available conformations.
Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and a strategy that prefers smaller conformations (denoted `P').
The `P' method selects such conformations where the deviation of the atomic distance from the geometric center are lowest, which leads 
to selection of the packed conformations (Fig.~\ref{m040c}).

The percentage of frames that were traversible by the ligands are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
name shows the number of selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denote the number of atoms of the ligand.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
For the ligand DCP, a higher traversability rates was achieved when 50 conformations are selected.
This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using less number of conformations or when preferring the packed conformations.

@@

The results are summarized in Tab.~\ref{tab::main}, which shows percentage of frames in which the tested planners \RA\ and \RB\ found
trajecotries close to the active site. 
The active site was considered as reached if the distance from the center of the ligand to the active site was less than 2~\AA.

The tables shows percentages of traversable MD snapshots only for the first tunnel, as no trajectories were found in the second tunnel
for most of the tested ligand.
The only except is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.

The proposed \RA\ planner outperforms the \RB\ planner in the number of frames where a solution was found.
The distances of the configuration trees (measured using 3D Euclidean metric) to the active site and the runtimes in each frame
are shown in Fig.~\ref{fig::comparison}.
The runtimes of both algorithms increases in those frames, where the tunnel is not traversable, i.e., where the distance
to the active site is more than 2~\AA (e.g., the frame number 60).

>>>>

%Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
%to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
%the random sample $\qrand$ is connected to the tree if the line segment  from $\qnear$ to $\qrand$ is collision-free. Otherwise, the retraction-step is performed.
%The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complexity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses 
the retractions only to the narrow passages.
Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on e.g. random vectors, obstacles vectors or medial axis in order to improve growth in the narrow passages.
In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed toward the medial axis.
%Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g. by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, RRT-ML~\cite{cortes2007mlrrt} is employed as the basic planner.
RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., that are responsible for conformation changes) if they hinder the growth of the tree.

@@

In this paper, we propose a modification of the RRT planner for computing trajectories through protein tunnels.
The input is a set of tunnels computed by Voronoi diagrams and the task is to find trajectories along the given tunnels
leading from outside the protein to the active site.
The tunnels are detected in a sequence of protein dynamics that is computed using MD simulations.

In contrast to~\cite{cortes2010simulating}, which finds any exit pathway, we compute the trajectories only along each tunnel separately.
This is motivated by the practical needs of chemists that are interested in the traversability of a selected tunnel.
Computing trajectories along a single tunnel also brings advantages from the motion planning point of view.
It decreases the volume of the configuration space to be searched  and increases the relative volume of the narrow passages.
To further cope with the narrow passage problem, the random samples are not generated in the whole configuration space, but only (mainly) around
the tunnel in a user-defined distance.
Therefore, the tunnels serve as a guiding path for the configuration tree~\cite{vonasek2009rrt}.

Due to the protein dynamics, the tunnel changes, move and merge, and their bottleneck changes in each frame.
It can therefore be expected, that the detected tunnels are rather narrow.
Based on the observation from MD simulations, the tunnels adapt to the ligands (and vice versa), so even narrow tunnels can serve as transportation path for thhe ligands TODO CITACE??.
To enable computing of trajectories in the narrow tunnels, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating,guieysse2008structure}.

The ligand flexibility is modeled using a predefined set of typical conformations, therefore the number of the dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
While classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
This metric measures distances between two configurations as the nearest distance of two atoms. 
This enables the metric to consider the shape of the ligand and it support retraction of the ligand to the narrow passages.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.

@@

%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The flexibility of a ligand is modeled using the set $\L$ of its conformations.
We assume that a library of all possible (in the considered resolution) conformations is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for motion planning.
%The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.

To enable motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
A discrete set of scales is used, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (the geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$ and the actual scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various conformations.
The traversability of the tunnel should therefore be evaluated using various starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation  parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

@@

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
%This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
This allows the tree to slightly detour from the tunnel, which improves performance around narrow tunnels.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
metric for the nearest-neighbor search.
The parameter $\rv$ influences the distribution of samples around the tunnel centerline. 
By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
of $\rv$ cause  exploration of paths around the tunnel.
We propose to set this parameter to the average tunnel width.

@@

The second proposed modification of RRT is the expansion procedure (Alg.~\ref{alg::expand}), which generates new collision-free nodes around $\qnear\in\T$.
For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qrand$ and tested for collisions.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
part $(x,y,z)$ is generated around $\qnear$.

@@

Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration
$q$ being tested.
This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}.
As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
The fast evaluation of this metric with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, can be achieved using KD-trees.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find nearest point towards $\qrand'$, which
is also the nearest neighbor of the ligand at $q$ to $\qrand$.

@@

The result of each planning trial is the tree $\T$ of collision-free configurations in which a path
between $\qinit$ (root of the tree) and $q'$ is found, where
 $q'$ is the nearest node to the end of the tunnel (measured using the 3D Euclidean metric). 
%The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
The path is found in the tree even if the tree does not approach $\qgoal$ close enough.
Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.

@@

In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
By using more conformations, there is a bigger chance that at least one of them can expand the tree towards the random sample $\qrand$.
On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are close to each other, but 
the atoms can also form a more longer shape (Fig.~\ref{fig::m040c}).

From motion planning point of view, both types of conformations are useful.
The `longer' conformations can better fit in a narrow passage, while the `packed' conformations can be better rotated at place 
To select the suitable conformations, a simple technique is proposed.  
For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
The conformations are sorted according to this deviation and then the desired number of conformations are selected from this sorted list uniformly.

\begin{figure}
\centering
%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
\includegraphics[width=0.15\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.12\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

The proposed approach was used to analyze tunnel traversability towards the active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
The molecular dynamics of the protein was computed using the Amber tool.
%100 frames of the dynamics were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.
%The second tunnel is more dynamic, i.e., it fluctuatest more

@@

Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), 
    denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
The Retraction-based RRT was implemented using 6D Euclidean metric.
The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configurations $\qnew$ that approaches $\qrand$.
The retraction step was realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.1$~\AA\ using 20 samples.
A collision-free sample that maximally approached $\qrand$ was selected and added to the tree, and the retraction continued from this sample.
The retraction step was repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per conformation, while \RA\ extends the tree only by one configuration.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for nearest-neighbor search and the time to solve these queries is significantly less than computing collision detection.
Parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is same.
For each conformation and tested scale, the \RA\ methods attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.

In the following experiments, we measure performance of the methods using the traversability rate, which is the percentage of the protein dynamic frames that are traversable.
A frame is considered traversable, if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) \times 10 trials approaches the active site to the distance less than 2~\AA.

@@

The results are shown in Tab.~\ref{tab::one} which shows traversability rate. 
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with bottleneck larger than the tested radius.
The row Planners shows traversable rate detected by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increases the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromised between the performance and the runtime.
The rates provided by \RA\ and \RB\ are similar in this case.

@@

The aim of this experiment is to verify the procedure to select ligand conformations for the motion planning.
The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations where available and for the ligand
m040 (1,5-Dichloropentane) with $7560$ available conformations.
Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and a strategy that prefers smaller conformations (denoted `P').
The `P' method selects such conformations where the deviation of the atomic distance from the geometric center are lowest, which leads 
to selection of the packed conformations (Fig.~\ref{fig::m040c}).

The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
name shows the number of selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denotes the number of atoms of the ligand.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
For the ligand DCP, higher traversability rates was achieved when 50 conformations were selected.
This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using less number of conformations or when preferring the packed conformations.

@@

The results are summarized in Tab.~\ref{tab::main}.
The tables shows percentages of traversable MD snapshots only for the first tunnel, as no trajectories were found in the second tunnel
for most of the tested ligand.
The only exception is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.

The proposed \RA\ planner provides higher traversability rate than \RB.
The distances of the configuration trees (measured using 3D Euclidean metric) to the active site and the runtimes in each frame
are shown in Fig.~\ref{fig::comparison}.
The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
to the active site is more than 2~\AA (e.g., the frame number 60).

The most time consuming part of the planners is the collision detection. 
Although both methods can utilize same amount of collision detection queries in the expansion step, \RB\ is slower than \RA.
The reason is, that \RA\ expands the trees more closer to the random samples $\qrand$ than \RB.
This results in faster exploration of the conifguration space and therefore, \RA\ needs less number of iterations to reach the goal
configuration.
We explain this behavior
